---
title: "Hands-on Exercise 9 (Part 4)"
description: "Visual Multivariate Analysis with Parallel Coordinates Plot"
author: "Quynh Trang Do"
date: "June 17, 2025"
date-modified: "last-modified"
format: html
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Install and launch necessary packages
```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

# Import data
```{r}
wh <- read_csv("data/WHData-2018.csv")
```

# Plotting Static Parallel Coordinates Plot
## Plotting a simple parallel coordinates
```{r}
ggparcoord(data = wh, 
           columns = c(7:12))
```

## Plotting a parallel coordinates with boxplot
```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

## Parallel coordinates with facet
```{r, fig.width=12}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```

## Rotating x-axis text label
```{r, fig.width=12}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))
```

## Adjusting the rotated x-axis text label
```{r, fig.width=15}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

# Plotting Interactive Parallel Coordinates Plot: parallelPlot methods
## The basic plot
```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```


## Rotate axis label
```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

## Changing the colour scheme
```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

## Parallel coordinates plot with histogram
```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```

