---
title: "Hands-on Exercise 9 (Part 2)"
description: "Visual Correlation Analysis"
author: "Quynh Trang Do"
date: "June 17, 2025"
date-modified: "last-modified"
format: html
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Install and launch necessary packages

```{r}
pacman::p_load(corrplot, ggstatsplot, tidyverse)
```

# Data preparation

## Import data

```{r}
wine <- read_csv("data/wine_quality.csv")
```

# Building Correlation Matrix: pairs() method

## Basic correlation matrix

```{r}
pairs(wine[,1:11])
```

```{r}
pairs(wine[,2:12])
```

## Draw the lower corner

### Lower-half

```{r}
pairs(wine[,2:12], upper.panel = NULL)
```

### Upper-half

```{r}
pairs(wine[,2:12], lower.panel = NULL)
```

## Include correlation coefficients

```{r}
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...) {
usr <- par("usr")
on.exit(par(usr))
par(usr = c(0, 1, 0, 1))
r <- abs(cor(x, y, use="complete.obs"))
txt <- format(c(r, 0.123456789), digits=digits)[1]
txt <- paste(prefix, txt, sep="")
if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
}

pairs(wine[,2:12], 
      upper.panel = panel.cor)
```

# Visualising Correlation Matrix: ggcormat()

## Basic plot

```{r, fig.width=10, fig,height=6}
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11)
```

## Add arguments

```{r, fig.width=10, fig.height=6}
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11,
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  title    = "Correlogram for wine dataset",
  subtitle = "Four pairs are no significant at p < 0.05"
)

ggplot.component = list(
    theme(text=element_text(size=5),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8)))
```

# Build multiple plots

```{r, fig.height=10, fig.width=15}
grouped_ggcorrmat(
  data = wine,
  cor.vars = 1:11,
  grouping.var = type,
  type = "robust",
  p.adjust.method = "holm",
  plotgrid.args = list(ncol = 2),
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  annotation.args = list(
    tag_levels = "a",
    title = "Correlogram for wine dataset",
    subtitle = "The measures are: alcohol, sulphates, fixed acidity, citric acid, chlorides, residual sugar, density, free sulfur dioxide and volatile acidity",
    caption = "Dataset: UCI Machine Learning Repository"
  )
)
```

# Visualising Correlation Matrix using corrplot Package

## Compute the correlation matrix

```{r}
wine.cor <- cor(wine[, 1:11])
```

## Plot

```{r}
corrplot(wine.cor)
```

## Visual geometrics

```{r}
corrplot(wine.cor, 
         method = "ellipse") 
```

## Layout

### Layout type: full, upper, or lower

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         type="lower")
```

### Turn off diagonal cells and change axis text label color

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         type="lower",
         diag = FALSE,
         tl.col = "black")
```

## Mixed layout

```{r, fig.width=10}
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

## Combine corrgram with the significant test

```{r}
# Compute p-value and confidence interval for each pair of variables
wine.sig = cor.mtest(wine.cor, conf.level= .95)

# Plot
corrplot(wine.cor,
         method = "number",
         type = "lower",
         diag = FALSE,
         tl.col = "black",
         tl.srt = 45,
         p.mat = wine.sig$p,
         sig.level = .05)
```

## Reorder a corrgram

```{r, fig.width=10}
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               order="AOE",
               tl.col = "black")
```

## Reorder a correlation matrix using hclust

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         tl.pos = "lt",
         tl.col = "black",
         order="hclust",
         hclust.method = "ward.D",
         addrect = 3)
```
