---
title: "Hands-on Exercise 9 (Part 3)"
description: "Heatmap for Visualising and Analysing Multivariate Data"
author: "Quynh Trang Do"
date: "June 17, 2025"
date-modified: "last-modified"
format: html
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Install and launch necessary packages
```{r}
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

# Data preparation
## Import data
```{r}
wh <- read_csv("data/WHData-2018.csv")
```

## Prepare data
```{r}
# change the rows by country name
row.names(wh) <- wh$Country

# transform wh data frame into a data matrix
wh1 <- dplyr::select(wh, c(3, 7:12))
wh_matrix <- data.matrix(wh)
```

# Static heatmap
```{r}
wh_heatmap <- heatmap(wh_matrix,
                      Rowv=NA, Colv=NA)
```

# Cluster heatmap
```{r}
wh_heatmap <- heatmap(wh_matrix)
```

## Normalize the matrix column-wise
```{r}
wh_heatmap <- heatmap(wh_matrix,
                      scale="column",
                      cexRow = 0.6, 
                      cexCol = 0.8,
                      margins = c(10, 4))
```

# Interactive heatmap
## Heatmaply
```{r}
heatmaply(mtcars)
```

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```

## Data tranformation
### Scaling method
```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)],
          scale = "column")
```

### Normalising method
```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]))
```

### Percentising method
```{r}
heatmaply(percentize(wh_matrix[, -c(1, 2, 4, 5)]))
```

## Clustering algorithm
### Manual approach
```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "ward.D")
```

### Statistical approach
**Step 1:** Use dend_expend() to determine the recommended clustering method to be used.
```{r}
wh_d <- dist(normalize(wh_matrix[, -c(1, 2, 4, 5)]), method = "euclidean")
dend_expend(wh_d)[[3]]
```

**Step 2:** Use find_k() to determine the optimal number of cluster.
```{r}
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)
```

**Step 3:** Plot the heatmap
```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "average",
          k_row = 3)
```

## Seriation
Use the seriation package to find an optimal ordering of rows and columns.
### Optimal Leaf Ordering (OLO) option
```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "OLO")
```

### Gruvaeus and Wainer (GW) option
```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "GW")
```

### Mean option
```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "mean")
```

### None option
```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none")
```

## Colour palettes
```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none",
          colors = Blues)
```

## Other features
```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          Colv=NA,
          seriate = "none",
          colors = Blues,
          k_row = 5,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="World Happiness Score and Variables by Country, 2018 \nDataTransformation using Normalise Method",
          xlab = "World Happiness Indicators",
          ylab = "World Countries"
          )
```


